Effects of the body force on the pedestrian and the evacuation dynamics

### 2020-03-04 paper.tex
quite el texto azul hasta corridor

The numerical \textcolor{blue}{\sout{setting} set} of the parameters may affect the dynamics of the 
pedestrians. Some \textcolor{blue}{\sout{settings} values}, however, yield similar collective dynamics.
We introduce \textcolor{blue}{dimensionless} magnitudes 
and proceed straightforward as indicated in \ref{appendix1}. We realize from 
\ref{appendix1} that only three (\textcolor{blue}{dimensionless}) parameters  are the true 
``control'' parameters for the collective dynamics. These are $\mathcal{A}$, 
$\mathcal{K}$, $\mathcal{K}_c$ as defined in \ref{appendix1}. Recall that 
$\mathcal{A}$ and $\mathcal{K}$ are precisely the same as in 
Ref.~\cite{dorso_2019}, but a novel $\mathcal{K}_c$ parameter has been 
introduced due to the body force.   \\

The logical relations between the \textcolor{blue}{dimensionless} parameters and the ``individual'' 
parameters can be studied by means of the Venn diagrams exhibited in 
Fig.~\ref{venn_diagram}. The parameters $\mathcal{A}$, $\mathcal{K}$ and 
$\mathcal{K}_c$ are represented as intersecting sets, and the ``individual'' 
parameters are represented as elements within each set. The shared elements 
between $\mathcal{A}$, $\mathcal{K}$ and $\mathcal{K}_c$ are placed inside the 
intersecting regions. \\

@@

\subfloat[\label{venn_2parameters}]
{\includegraphics[width=0.49\columnwidth]{./venn_2parameters.png}}
\hfill    
\subfloat[\label{venn_3parameters}]
{\includegraphics[width=0.49\columnwidth]{./venn_3parameters.png}
}\\
\caption[width=0.47\columnwidth]{Venn diagrams for the \textcolor{blue}{dimensionless} parameters
appearing in the equation of motion (\ref{eqn_motion}) (see 
Appendix~\ref{appendix1} for details). The sets correspond to 
$\mathcal{A}=\{\tau/m,1/v_d,A\}$, $\mathcal{K}=\{\tau/m,B,k_t\}$ and 
$\mathcal{K}_c=\{\tau/m,1/v_d,B,k_n\}$. (a) The Venn diagram representation if no 
body force is introduced in the SFM (only sets $\mathcal{A}$ and $\mathcal{K}$, 
as in Ref.~\cite{dorso_2019}). (b) The Venn diagram representation for the sets 
$\mathcal{A}$, $\mathcal{K}$ and $\mathcal{K}_c$. }
\label{venn_diagram}
\end{figure}

@@

Fig.~\ref{venn_2parameters} represents the situation when $\mathcal{K}_c$ is 
absent. Notice that $A/v_d$ or $k_t B$ may control the collective 
dynamic, in spite of the ``assertive'' attitude. The individual character of 
$v_d$ or $B$ appears somehow ``loosely'' in the crowd dynamic. We mean by 
``loosely'' that any numerical \textcolor{blue}{\sout{setting} set for} these parameters may be 
counterbalanced by the right \textcolor{blue}{\sout{setting} choice } of $A$ or $k_t$, keeping the 
colletive dynamic (qualitatively) unchanged.  \\

@@

\begin{itemize}
 \item[(a)] The ``faster-is-slower'' (or the ``faster-is-faster'') phenomenon 
occurs when varying the values of $v_d$ (see Section~\ref{bottleneck}). This is 
equivalent to vary simultaneously $\mathcal{A}$ and $\mathcal{K}_c$ by the same 
amount in the (\textcolor{blue}{dimensionless}) parameter space (see Section~\ref{parameters}). We 
may visualize this sampling procedure as moving along a straight line in the 
space $(\mathcal{A},\mathcal{K},\mathcal{K}_c)$. Further variations of 
$k_n$ turns the sample points out of this line, but on the plane of constant
$\mathcal{K}_c$. We will step up $k_n$ by several orders of magnitude in order 
to get the big picture of this surface.

\item[(b)] We will consider the parameter \textcolor{blue}{set from} Ref.~\cite{helbing_2000} 
as a starting point for exploring the parameter space. The reason for this is 
that a complete set of experimental parameters is still not available (to our 
konwledge). We are aware, though, of the drawbacks \textcolor{blue}{of} this choice, say, the 
unrealistic meanings for the length $B$ and the compressibility $k_n$ (see 
Section~\ref{experimental}). But these are irrelevant in the context of the 
(\textcolor{blue}{dimensionless}) parameter space and the corresponding collective dynamics (see 
Section~\ref{introduction}).

@@

In order to keep the model as simple as possible, we will only consider 
\textcolor{blue}{isotropic pedestrian interactions}, since \textcolor{blue}{anisotropic
interactions do} not play a relevant role at high densities. \\

@@

The implemented SFM parameters were the same as those in 
Ref.~\cite{helbing_2000} (at the beginning of the exploratory procedure only). 
But the pedestrian's mass and radius were set to the more realistic values of 
$70\,$kg and $0.23\,$m, respectively. The force interactions between 
pedestrians were limited, however, to a cut-off distance of $0.88\,$m for 
attaining a \textit{privacy sphere} \textcolor{blue}{that excludes second neighbors}. The desired 
velcity was always set to $1\,$m/s in the corridor situation. Besides, the 
explored values of $v_d$ for the bottleneck scenario ranged from $1\,$m/s to the 
extremely anxious situation of $10\,$m/s. \\

@@

Fig.~\ref{vd_vs_te} shows the evacuation time as a function of the pedestrian's 
desired velocity for different values of $k_n$. The evacuation 
time is defined as the time \textcolor{blue}{lapsed} until 80\% of the pedestrians have 
left the room. In this section we will focus on the evacuation time for 
$2\,$m/s$<v_d<10\,$m/s.\\

@@

Three behavioral patterns can be distinguished in Fig.~\ref{vd_vs_te}. 
Each pattern can \textcolor{blue}{display} a positive slope, a negative slope
or both. The interval in which the slope is positive means that the 
harder the pedestrians try to get out (higher $v_d$), the 
longer it takes them to evacuate. This is the Faster-is-slower 
(FIS) regime. Conversely, the interval in which the slope is 
negative  corresponds to a Faster-is-Faster (FIF) regime (the 
harder they try, the quicker they leave).\\

@@

Despite the presence of the FIS or FIF pattern, the evacuation time at a 
fixed value of $v_d$ decreases for increasing values of $k_n$ (\textcolor{blue}{within} the examined interval). 
This means that stiffer pedestrians evacuate faster than soft pedestrians. 
To further investigate this 
phenomenon, we computed the mean velocity of the whole crowd as a function of the 
stiffness $k_n$. Fig.~\ref{kn_vs_vx_bottleneck} shows the actual mean velocity in the 
$x-$direction as a function of the stiffness $k_n$ for three different desired 
velocities. In all cases, the velocity increases with the stiffness. The most significative 
increment of velocity is in the interval \textcolor{blue}{$k_n>\,10^{4}$}. \\

@@

Fig.~\ref{degree_vd} shows the mean degree of the contact network as a function  
of the desired velocity. The degree of a node is defined as the number of links 
that \textcolor{blue}{connects} this node to any other node. This means, the number of pedestrians 
that are in physical contact with a given pedestrian. The mean degree is the 
average of the degree over all the nodes (pedestrians) and over the whole 
sampled interval. We computed mean values only after the system reached the stationary 
state, that is, after a well-formed bulk has been established.\\

Notice that the mean degree increases as $v_d$ increases, as expected. This expresses the fact that
higher $v_d$ values accomplish higher densities, \textcolor{blue}{forcing} individuals to touch each 
other. For a given $v_d$, the mean degree reduces as the $k_n$ 
value increases. A noticeable decrease in the mean degree can be seen for the 
highest explored value of $k_n$. This \textcolor{blue}{\sout{was not expected and}} opens the question on how would
this affect the sliding friction among pedestrians.\\

@@

The regime for $v_d<2\,$m/s is qualitatively different from the clogged regime since most of the 
pedestrians are not touching each other (as was already observed in Ref~\cite{dorso_2005,dorso_2011}). The only pedestrians who touch each 
other are the ones who are being re-injected on the opposite side of the room.
\textcolor{blue}{These collide with the bulk}, providing some kind of overlap during very short time. We will
not analize this regime.\\

@@

The four snapshots in Fig.~\ref{network_bottleneck} confirm (visually) the fact that more rigid pedestrians \textcolor{blue}{ease} the crowded environment, widening the occupied region. At $v_d=10\,$m/s (the maximum explored velocity), it can hardly be found pedestrians with degree 6 when $k_n=1.2\,$E6, while a lot of them are present for $k_n$=0.\\

@@

We may summarize this Section as follows. We explored the parameter space along $v_d$ and $k_n$ for the bottleneck situation. This is similar to explore the \textcolor{blue}{dimensionless} plane $(\mathcal{A},\mathcal{K}_c)$. Soft pedestrians attain a FIS or FIS+FIF evacuation pattern but, stiff pedestrians (for any fixed $\mathcal{A}$ value within the explored range) exhibit a single FIF pattern. Stiffness affects either the presence of the blocking clusters and the pedestrians overlap. The less they overlap, the less intense becomes the sliding friction among them. Additionally, the fewer the blocking clusters the easier they get out. \\

>>>>

The numerical of the parameters may affect the dynamics of the 
pedestrians. Some values, however, yield similar collective dynamics.
We introduce dimensionless magnitudes 
and proceed straightforward as indicated in \ref{appendix1}. We realize from 
\ref{appendix1} that only three (dimensionless) parameters  are the true 
``control'' parameters for the collective dynamics. These are $\mathcal{A}$, 
$\mathcal{K}$, $\mathcal{K}_c$ as defined in \ref{appendix1}. Recall that 
$\mathcal{A}$ and $\mathcal{K}$ are precisely the same as in 
Ref.~\cite{dorso_2019}, but a novel $\mathcal{K}_c$ parameter has been 
introduced due to the body force.   \\

The logical relations between the dimensionless parameters and the ``individual'' 
parameters can be studied by means of the Venn diagrams exhibited in 
Fig.~\ref{venn_diagram}. The parameters $\mathcal{A}$, $\mathcal{K}$ and 
$\mathcal{K}_c$ are represented as intersecting sets, and the ``individual'' 
parameters are represented as elements within each set. The shared elements 
between $\mathcal{A}$, $\mathcal{K}$ and $\mathcal{K}_c$ are placed inside the 
intersecting regions. \\

@@

\subfloat[\label{venn_2parameters}]
{\includegraphics[width=0.49\columnwidth]{./venn_2parameters.png}}
\hfill    
\subfloat[\label{venn_3parameters}]
{\includegraphics[width=0.49\columnwidth]{./venn_3parameters.png}
}\\
\caption[width=0.47\columnwidth]{Venn diagrams for the dimensionless parameters
appearing in the equation of motion (\ref{eqn_motion}) (see 
Appendix~\ref{appendix1} for details). The sets correspond to 
$\mathcal{A}=\{\tau/m,1/v_d,A\}$, $\mathcal{K}=\{\tau/m,B,k_t\}$ and 
$\mathcal{K}_c=\{\tau/m,1/v_d,B,k_n\}$. (a) The Venn diagram representation if no 
body force is introduced in the SFM (only sets $\mathcal{A}$ and $\mathcal{K}$, 
as in Ref.~\cite{dorso_2019}). (b) The Venn diagram representation for the sets 
$\mathcal{A}$, $\mathcal{K}$ and $\mathcal{K}_c$. }
\label{venn_diagram}
\end{figure}

@@

Fig.~\ref{venn_2parameters} represents the situation when $\mathcal{K}_c$ is 
absent. Notice that $A/v_d$ or $k_t B$ may control the collective 
dynamic, in spite of the ``assertive'' attitude. The individual character of 
$v_d$ or $B$ appears somehow ``loosely'' in the crowd dynamic. We mean by 
``loosely'' that any numerical set for these parameters may be 
counterbalanced by the right choice of $A$ or $k_t$, keeping the 
collective dynamic (qualitatively) unchanged.  \\

@@

\begin{itemize}
 \item[(a)] The ``faster-is-slower'' (or the ``faster-is-faster'') phenomenon 
occurs when varying the values of $v_d$ (see Section~\ref{bottleneck}). This is 
equivalent to vary simultaneously $\mathcal{A}$ and $\mathcal{K}_c$ by the same 
amount in the (dimensionless) parameter space (see Section~\ref{parameters}). We 
may visualize this sampling procedure as moving along a straight line in the 
space $(\mathcal{A},\mathcal{K},\mathcal{K}_c)$. Further variations of 
$k_n$ turns the sample points out of this line, but on the plane of constant
$\mathcal{K}_c$. We will step up $k_n$ by several orders of magnitude in order 
to get the big picture of this surface.

\item[(b)] We will consider the parameter set from Ref.~\cite{helbing_2000} 
as a starting point for exploring the parameter space. The reason for this is 
that a complete set of experimental parameters is still not available (to our 
konwledge). We are aware, though, of the drawbacks of this choice, say, the 
unrealistic meanings for the length $B$ and the compressibility $k_n$ (see 
Section~\ref{experimental}). But these are irrelevant in the context of the 
(dimensionless) parameter space and the corresponding collective dynamics (see 
Section~\ref{introduction}).

@@

In order to keep the model as simple as possible, we will only consider 
isotropic pedestrian interactions, since anisotropic
interactions do not play a relevant role at high densities. \\

@@

The implemented SFM parameters were the same as those in 
Ref.~\cite{helbing_2000} (at the beginning of the exploratory procedure only). 
But the pedestrian's mass and radius were set to the more realistic values of 
$70\,$kg and $0.23\,$m, respectively. The force interactions between 
pedestrians were limited, however, to a cut-off distance of $0.88\,$m for 
attaining a \textit{privacy sphere} that excludes second neighbors. The desired 
velcity was always set to $1\,$m/s in the corridor situation. Besides, the 
explored values of $v_d$ for the bottleneck scenario ranged from $1\,$m/s to the 
extremely anxious situation of $10\,$m/s. \\

@@

Fig.~\ref{vd_vs_te} shows the evacuation time as a function of the pedestrian's 
desired velocity for different values of $k_n$. The evacuation 
time is defined as the time lapsed until 80\% of the pedestrians have 
left the room. In this section we will focus on the evacuation time for 
$2\,$m/s$<v_d<10\,$m/s.\\

@@

Three behavioral patterns can be distinguished in Fig.~\ref{vd_vs_te}. 
Each pattern can display a positive slope, a negative slope
or both. The interval in which the slope is positive means that the 
harder the pedestrians try to get out (higher $v_d$), the 
longer it takes them to evacuate. This is the Faster-is-slower 
(FIS) regime. Conversely, the interval in which the slope is 
negative  corresponds to a Faster-is-Faster (FIF) regime (the 
harder they try, the quicker they leave).\\

@@

Despite the presence of the FIS or FIF pattern, the evacuation time at a 
fixed value of $v_d$ decreases for increasing values of $k_n$ (within the examined interval). 
This means that stiffer pedestrians evacuate faster than soft pedestrians. 
To further investigate this 
phenomenon, we computed the mean velocity of the whole crowd as a function of the 
stiffness $k_n$. Fig.~\ref{kn_vs_vx_bottleneck} shows the actual mean velocity in the 
$x-$direction as a function of the stiffness $k_n$ for three different desired 
velocities. In all cases, the velocity increases with the stiffness. The most significative 
increment of velocity is in the interval $k_n>\,10^{4}$. \\

@@

Fig.~\ref{degree_vd} shows the mean degree of the contact network as a function  
of the desired velocity. The degree of a node is defined as the number of links 
that connects this node to any other node. This means, the number of pedestrians 
that are in physical contact with a given pedestrian. The mean degree is the 
average of the degree over all the nodes (pedestrians) and over the whole 
sampled interval. We computed mean values only after the system reached the stationary 
state, that is, after a well-formed bulk has been established.\\

Notice that the mean degree increases as $v_d$ increases, as expected. This expresses the fact that
higher $v_d$ values accomplish higher densities, forcing individuals to touch each 
other. For a given $v_d$, the mean degree reduces as the $k_n$ 
value increases. A noticeable decrease in the mean degree can be seen for the 
highest explored value of $k_n$. This opens the question on how would
this affect the sliding friction among pedestrians.\\

@@

The regime for $v_d<2\,$m/s is qualitatively different from the clogged regime since most of the 
pedestrians are not touching each other (as was already observed in Ref~\cite{dorso_2005,dorso_2011}). The only pedestrians who touch each 
other are the ones who are being re-injected on the opposite side of the room.
These collide with the bulk, providing some kind of overlap during very short time. We will
not analize this regime.\\

@@

The four snapshots in Fig.~\ref{network_bottleneck} confirm (visually) the fact that more rigid pedestrians ease the crowded environment, widening the occupied region. At $v_d=10\,$m/s (the maximum explored velocity), it can hardly be found pedestrians with degree 6 when $k_n=1.2\,$E6, while a lot of them are present for $k_n$=0.\\

@@

We may summarize this Section as follows. We explored the parameter space along $v_d$ and $k_n$ for the bottleneck situation. This is similar to explore the dimensionless plane $(\mathcal{A},\mathcal{K}_c)$. Soft pedestrians attain a FIS or FIS+FIF evacuation pattern but, stiff pedestrians (for any fixed $\mathcal{A}$ value within the explored range) exhibit a single FIF pattern. Stiffness affects either the presence of the blocking clusters and the pedestrians overlap. The less they overlap, the less intense becomes the sliding friction among them. Additionally, the fewer the blocking clusters the easier they get out. \\
